European or American? A re-examination of the transatlantic influence over the IMF

### 2025-07-31 manuscript/manuscript.qmd
remove a para
Add abstract draft

---
title: European or American? A re-examination of the transatlantic influence over the IMF
date: last-modified
authors:
  - name: Dianyi Yang
    email: d.yang17@lse.ac.uk
    orcid: 0009-0004-4652-3429
    correspondence: true
  - name: Tong Hu
    orcid: 0009-0003-5234-8164
  - name: Zhenyi Chen
    orcid: 0009-0008-2910-5187
format:
  #native: default
  #docx: default
  orcid-pdf:
    keep-tex: true
    number-sections: true
    include-in-header:
      - text: |
          \newcommand{\centerstart}{\begin{center}}
          \newcommand{\centerend}{\end{center}}
          \usepackage{amsthm, thm-restate}
          \newtheorem{hyp}{Hypothesis}
          \newtheorem{subhyp}{Hypothesis}[hyp]
          \renewcommand{\thesubhyp}{\thehyp\alph{subhyp}}
          \usepackage{chngcntr}
          \usepackage{apptools}
          \AtAppendix{\counterwithin{figure}{section}}
abstract: This is an abstract.
keywords: IMF, US hegemony, Europe
execute:
  echo: false
  warning: false
  error: false
  message: false
---

@@

Compared to the extensive literature on US dominance in the IMF, scholarship on European influence is relatively limited and often fragmented. Many studies underscore that EU member states are institutionally dispersed across multiple constituencies, hindering coherent collective action [@smaghi2004; @miescu2014; @thacker1999]. Still, other works argue that European countries wield significant structural advantages—including leadership appointments, voting power, and coordination mechanisms—that enable them to shape lending outcomes in line with their interests [@copelovitch2010a; @keating2024; @lutz2014; @stone2004].

European influence reflects a strategic combination of formal power and coordinated action. EU countries collectively control over 30% of IMF voting rights, and if they were to unify, they could potentially counterbalance US veto power [@brandner2009a; @leech2005a; @smaghi2004]. This is reinforced by the long-standing transatlantic norm allocating IMF leadership to a European national [@kahler2001]. Through platforms such as EURIMF and SCIMF, EU members have harmonised policy positions and influenced loan design. During the Eurozone crisis, France and Germany pushed for IMF involvement to stabilise the region while embedding austerity-oriented conditionality that protected their banking systems [@hodson2014; @kickert2019a; @veron2016]. Lending terms were shaped to reflect creditor priorities, with evidence that conditionality was both intensified or relaxed based on geopolitical alignment [@cain2022; @dreher2015; @lutz2014]. Beyond Europe, the UK and France have used IMF concessional lending in Africa to maintain post-colonial ties and commercial access [@moyo2024; @stone2004].

Despite Europe’s institutional advantages, scholars remain skeptical of its actual effectiveness in the IMF. Internal divergences—such as France’s growth-oriented stance versus Germany’s fiscal conservatism—combined with fragmented constituencies, often impede coherent action and reduce policy influence [@miescu2014; @dreher2015; @copelovitch2010a]. Moreover, Europe’s disproportionate voting power relative to its population raises legitimacy concerns and is seen as disadvantaging developing countries [@woodward2024; @leech2005a]. While proposals to consolidate EU seats aim to improve efficiency and counterbalance US dominance, critics warn such reforms may merely reinforce dominance by a few large European states rather than broaden representation [@leech2005a; @copelovitch2010a].

While several studies examine European involvement in IMF operations, most focus on individual cases or regional crises, lacking generalisable conclusions. Existing literature often treats Europe as a secondary actor or subsumes it under undifferentiated “Western” influence, thereby overlooking intra-Western variation. By analytically separating US and European roles, our study addresses these gaps and offers a comparative framework for understanding divergent patterns of influence within global financial governance.

>>>>

---
title: European or American? A re-examination of the transatlantic influence over the IMF
date: last-modified
authors:
  - name: Dianyi Yang
    email: d.yang17@lse.ac.uk
    orcid: 0009-0004-4652-3429
    correspondence: true
  - name: Tong Hu
    orcid: 0009-0003-5234-8164
format:
  #native: default
  #docx: default
  orcid-pdf:
    keep-tex: true
    number-sections: true
    include-in-header:
      - text: |
          \newcommand{\centerstart}{\begin{center}}
          \newcommand{\centerend}{\end{center}}
          \usepackage{amsthm, thm-restate}
          \newtheorem{hyp}{Hypothesis}
          \newtheorem{subhyp}{Hypothesis}[hyp]
          \renewcommand{\thesubhyp}{\thehyp\alph{subhyp}}
          \usepackage{chngcntr}
          \usepackage{apptools}
          \AtAppendix{\counterwithin{figure}{section}}
abstract: "This paper investigates whether the United States or Europe holds greater influence over the International Monetary Fund (IMF)—a longstanding yet underexamined question in the study of global economic governance. While the Fund is widely regarded as dominated by “Western” powers, most existing research either focuses exclusively on American influence or treats the West as a unified bloc, masking important variation between its two key actors. Drawing on panel data from 1980 to 2010, we construct separate measures of US and European influence using principal component analysis of commonly used alignment indicators—UN voting patterns, bilateral trade, and banking exposure—and estimate a series of Tobit and Probit models to assess their impact on four key IMF lending outcomes: loan-to-GDP ratios, participation rates, loan approvals, and conditionality. We find that European influence is strongly associated with higher loan volumes, more frequent participation, and greater likelihood of loan approval, while US influence is not. However, American ties are significantly correlated with fewer conditions attached to loans. Additional analysis shows that European influence is not driven by regional or colonial favouritism: neither European nor African borrowers benefit more from European ties—and may, in fact, fare worse—than other countries with comparable alignment. These findings complicate existing assumptions about IMF decision-making, highlighting the need to distinguish between types of Western influence. They also suggest that Europe wields broader institutional influence over access to IMF resources, while US power manifests more selectively through conditionality. In doing so, the paper contributes to debates on hegemony, institutional legitimacy, and intra-Western dynamics in global economic governance."
keywords: IMF, US hegemony, Europe
execute:
  echo: false
  warning: false
  error: false
  message: false
---

@@

In comparison, the literature on European influence is notably smaller. The most visible manifestation of European power within the Fund is the tradition of European leadership: all Managing Directors of the IMF since its inception have been European nationals [@kahler2001]. Scholars have also pointed to Europe-backed lending decisions, such as those during the Eurozone crisis (notably Greece) [@lipscy2018; @hodson2014], and loans to former colonies in Africa [@stone2004]. However, while these studies offer important insights, they tend to remain case-specific or anecdotal. Crucially, they stop short of offering a systematic comparison between American and European influence, leaving open the question of how these two centres of power operate within the IMF.
